The evolution of gene expression in the term placenta of viviparous mammals

### 2016-12-01 presentation/placenta_viviparous_presentation.Rnw
use the new shapes and core without metatherians

\begin{frame}{Core Placenta Transcriptome}
\begin{figure}
<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
      + geom_raster(aes(fill=log10(FPKM)))
      + scale_x_discrete(expand=c(0,0))
      + scale_y_discrete(expand=c(0,0))
      + scale_fill_gradient(low = grad.def[["low"]],
                            high = grad.def[["high"]],
                            space = grad.def[["space"]],
                            na.value = grad.def[["na.value"]],
                            limits=c(0,4),
                            guide=guide_colorbar(title=expression(log[10](FPKM)))
                            )
      + ylab("Human Ortholog Name")
      + xlab("Species")
      + theme_grey()
      + theme(legend.position="top",
              axis.ticks = element_blank(),
              axis.text.y = element_text(face="italic"),
              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
      )
@ 
\end{figure}
\end{frame}

@@

cpt.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species," species"),
              x
              )
    nc.levels <- paste0(num.species," of ",num.species," species")
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species," species"),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species," species"))
    }
    return(factor(x,levels=nc.levels))
}

@@

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Minimum Expression in`:=
                                fix.expression.names(`Minimum Expression in`)]

setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")

print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
             aes(x=`Expression (FPKM)`,color=`Minimum Expression in`,y=num_genes))
      + geom_line(lwd=1)
      + xlab("Minimum Expression cutoff (FPKM)")
      + ylab("Genes with N-th Minimum Expression")
      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + geom_vline(xintercept=10)
      + geom_segment(mapping=aes(y=V1,
                         yend=V1,
                         xend=10,
                         x=1000,
                         color=`Minimum Expression in`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Minimum Expression in`],
                     arrow=arrow(length=unit(0.5,"cm"))
                   )
      + geom_text(mapping=aes(x=1000,
                      y=V1,
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Minimum Expression in`],
                  show.legend=FALSE,
                  hjust=0
                  )
      + guides(color=guide_legend(title="Minimum\nExpression in"))
      )
@ 
\end{figure}
\end{frame}

>>>>

\begin{frame}{Core Placenta Transcriptome}
\begin{figure}
<<core.placenta.transcriptome,fig.width=7,fig.height=7*9/16,out.height="0.8\\textheight",echo=FALSE>>=
cptl <-
    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                            Species) & housekeeping==FALSE,]
cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
                                          capfirst(names(species.names))))]
cptl[,human_name:=factor(human_name,rev(levels(as.factor(human_name))))]

print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual("Species",
                         guide=guide_legend(ncol=3,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      # scale_color_discrete(NULL)+
      # coord_flip()+
      xlab("Human Ortholog Name")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      theme(text=element_text(size=9),
            legend.position=c(0.99,0.99),
            axis.text.x=element_text(face="italic",size=7,angle=90,hjust=1,vjust=0.5),
            legend.margin=unit(0,"mm"),
            legend.justification=c(1,1),
            plot.margin=unit(c(0,0,0,0),"mm")
            ))
@ 
\end{figure}
\end{frame}

@@

cpt.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression",
             "all_but_metatherian_expression"
             ),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Min. Exp. in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_but_metatherian_expression$",
              paste0("All ",num.species-1," Eutherians"),
              x
              )
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species," Species"),
              x
              )
    nc.levels <- c(paste0("All ",num.species-1," Eutherians"),
                   paste0(num.species," of ",num.species," Species")
                   )
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species," Species"),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species," Species"))
    }
    return(factor(x,levels=nc.levels))
}

@@

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Min. Exp. in`:=
                                fix.expression.names(`Min. Exp. in`)]
setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")

print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
             aes(x=`Expression (FPKM)`,color=`Min. Exp. in`,y=num_genes))
      + geom_line(lwd=1)
      + xlab("Minimum Expression cutoff (FPKM)")
      + ylab("Number of Genes Meeting FPKM Cutoff")
      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + geom_vline(xintercept=10,linetype="dotted")
      + geom_segment(mapping=aes(y=V1-ifelse(V1<80,20,0),
                         yend=V1,
                         xend=10,
                         x=sapply(V1,
                                  function(x){
                                      cpt.distribution.shape[num_genes > x,
                                                             max(`Expression (FPKM)`)*4]
                                  }),
                         color=`Min. Exp. in`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Min. Exp. in`],
                     arrow=arrow(length=unit(0.5,"cm")),
                     show.legend=FALSE
                     )
      + geom_text(mapping=aes(x=sapply(V1,
                                       function(x){
                                           cpt.distribution.shape[num_genes > x,
                                                                  max(`Expression (FPKM)`)*4.1]
                                       }),
                      y=V1-ifelse(V1<80,20,0),
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Min. Exp. in`],
                  show.legend=FALSE,
                  hjust=0,
                  size=8*(5/14)
                  )
      + geom_segment(mapping=aes(yend=V1,
                         y=100,
                         xend=1,
                         x=1,
                         color=`Min. Exp. in`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 1 &
                                                 `Min. Exp. in` == 'All 13 Eutherians',
                                                 max(num_genes),`Min. Exp. in`],
                     arrow=arrow(length=unit(0.5,"cm")),
                     show.legend=FALSE
                     )
      + geom_text(mapping=aes(x=1,
                      y=100,
                      label=paste0(V1," genes"),
                      override=TRUE),
                  angle=90,
                  data=cpt.distribution.shape[`Expression (FPKM)` > 1 &
                                              `Min. Exp. in` == 'All 13 Eutherians',
                                              max(num_genes),`Min. Exp. in`],
                  show.legend=FALSE,
                  hjust=1,
                  size=8*(5/14)
                  )
      + guides(color=guide_legend(title=paste0("Minimum\nExpression in"),
                                  keywidth=unit(3,"mm"),
                                  keyheight=unit(3,"mm"),
                                  title.hjust=0.5,
                                  label.position="right",
                                  label.theme=element_text(size=7,angle=0)
                                  ))
      + theme_bw()
      + theme(text=element_text(size=9),
              legend.position=c(0.99,0.99),
              legend.margin=unit(0,"mm"),
              legend.justification=c(1,1),
              plot.margin=unit(c(0,0,0,0),"mm")
              )
      )
@ 
\end{figure}
\end{frame}
